Structural replication of nanoporous media using procedural noise

### 2018-05-29 ms.tex
Fixed errors and tex structure

These physically based models may not always include all relevant effects, can be quite computationally expensive when used to generate many samples, and cannot easily be extended beyond the initial size of the simulated systems. 
For example, to prepare a simulation of nanoporous silica, we need to carry out a sequence of large scale simulations in molecular dynamics, which is time consuming, computationally expensive and the results canno be exteded beyond the intial size of the simulated systems without performing another simulation \cite{Adarsh}. 
On the other hand, the statistical models often do not include all the correlations induced by physical processes or have limited number of tunable parameters. 
In particular, statistical models have challenges in reproducing atomic-scale structures of nanoporous media.

Meanwhile, there has been a recent growth in advanced random noise models that open for new ways to generate structured noise with statistical similarities to real samples, such as the use of Perlin noise. 
Such noise models have been used to model systems with complex statistical correlations such as in GAMER\cite{groeneboom:2014}, and show great promise in being adaptable to many different physical systems. 
Also, such noise models are continuous and extendable, so that arbitrary large systems or systems with arbitrary high precision can be modeled. 
While these models show great promise in many fields \cite{}, they have not yet been systematically applied to characterize or generate nanoporous media with statistical properties corresponding to that of real systems.

@@

Another example, widely used in contemporary game content generation, is having prefabs, or pre-fabricated 3D models such as doors, walls, tiles and roof segments automatically being stitched together to create buildings.
When designing a city, instead of creating each building manually, you only need specify the position, area and height of a structure.
Then, by using an algorithm that combines these prefabs to render the building, it is possible to create cities of arbitrary size with a single command.
Naturally, it is also possible to create procedural methods that design the layout and placement of these buildings.
See figure \ref{fig:orphancity} for an example of such a city generator tool, developed by one of the authors of this paper.

@@

An overview of the current state of procedural noise functions can be
found in \cite{lagae:2010}. Perlin noise is a band limited repeatable pseudo-random function
from $\mathbb R^n \to \mathbb R$ that is computationally fast, and has
properties which make it suitable for creating patterns that mimic
nature. Perlin noise is an approximation to Gaussian filtered noise
implemented as a pseudo-random spline. The algorithm uses a predefined
grid of gradients pointing in a random direction, where for any point
within a grid, we interpolate the gradients from the four corners. The
algorithm is as follows:
\begin{itemize}
  \item[1.] For a point, obtain the four closest gradients in the grid.
   \item [2.] For each of the gradients, calculate the dot product
     between the gradient and a vector defined by the  the distance
     between the the point and the current grid corner. 
   \item [3.] Instead of linear interpolation $F(t) = t$, use a fade function to produce smoother interpolated values. Perlin noise typically uses the following fade function: $F(t) = t ^3 (t  (6 - 15) + 10)$.
    \item [4.] Obtain the final value by linearly interpolating between the x-axis and
      use this result to linearly interpolate the y-axis. 
\end{itemize} 
Simplex noise works in a similar manner, but works on on simplices instead of hypergrids, and so requires fewer computational steps, scales better in higher ($\ge 3$) dimensions and has fewer anisotropic properties.

@@

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\Omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\Omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In one dimension, we show equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern is depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use simplex over regular Perlin noise as the main noise function throughout this paper.

@@

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce nanoporous structures that are statistically similar to physical simulations of porous media.
For a given data set and a model, we wish to estimate the parameters that best represent the data. 
This is performed through a full likelihood analysis in order to obtain the posterior probability distribution of the parameters. 
The model measure of the analysis is the radial distribution function $g(r)$, for which the parameters will be maximised to fit. 
In addition, we also investigate the porosity and surface area of the system. 
The simulated system is confined to a cube of size $\SI{159} {\angstrom}^3$, but since the procedural noise function is continuous on all scales, the box could be of arbitrary size.

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image: low persistence (large scale dominates), middle image: medium persistence (small scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}
We construct a generic sum over procedural noise octaves with the expression: 
\begin{equation}
  M_1(\bar \theta) = M_1(\vec x, \psi, \Omega, \nu) = \sum_{\omega=0}^{\Omega} \psi^{\omega -1}   N\big((\vec x + \sigma \cdot \omega)\nu \cdot 2^{\omega-1} \big),
\label{eq:noisemodel1}
\end{equation}
where $\Omega$ is the total amount of octaves, $\psi$ the persistence (scale falloff), $N$ a generic $N$-dimensional noise function, $\sigma$ a predefined random vector that removes anisotropic artefacts and $\nu$ the frequency modifier representing the initial largest scales in the model. For simplicity, $\bar \theta$ here denotes the set of model parameters. For each mode, the frequency is thus multiplied, while the overall amplitude is damped by a given persistence. In the end, we simulate a large set of bulk particles without any walls, and use the noise function to remove particles that are not within a procedural void given by a threshold $\tau$. Three examples of this model are depicted in figure \ref{fig:model_example}, where we have used $\Omega=6$ octaves and the scale is $\nu=0.01$ and cut-off $\tau=0$. The left image depicts low persistence ($\psi = 0.0$), where only the first octave dominates (corresponding to large scales). The middle picture shows the same parameters with a higher persistence $(\psi = 0.5)$, where smaller scales become more dominating. In the rightmost picture, the effect of the cut-off threshold $\tau = -0.3$ is shown to emphasize threshold cut-off.

@@

\begin{figure*}
\includegraphics[width=.99\textwidth]{mock_data_results_pts.png}
\caption{Results from the model validation simulation. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Note that all input parameters were reproduced well within $1 \sigma$ (inner contour line).}
\label{fig:mockdataresults_pts}
\end{figure*}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{gofr_figure.pdf}
\caption{$g(r)$ for three data sets: the simulated SiO$_2$ input data set (red), the best-fit parameter model (blue) and a random low-fit parameter model (green).}
\label{fig:gofr1}
\end{figure}
Finally, we depict a comparison of the actual simulated input SiO$_2$ data set together with a representation of the best-fit parameters in figure \ref{fig:porous_vs_model}   
\begin{figure}
\includegraphics[width=.45\textwidth]{comparison.png}
\caption{Right side: A realisation of the best-fit parameters $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$. Left side: Simulated SiO$_2$ input data set used for parameter estimation. Although the pore geometry seems to be similar, we see some artifacts, such as single loose atoms in the generated geometry.}
\label{fig:porous_vs_model}
\end{figure}

>>>>

These physically based models may not always include all relevant effects, can be quite
computationally expensive when used to generate many samples, and cannot easily be extended
beyond the initial size of the simulated systems. 
For example, to prepare a simulation of nanoporous silica, we need to carry out a sequence
of large scale simulations in molecular dynamics, which is time consuming, computationally
expensive and the results cannot be exteded beyond the intial size of the simulated systems
without performing another simulation \cite{Adarsh}. 
On the other hand, the statistical models often do not include all the correlations induced
by physical processes or have limited number of tunable parameters. 
In particular, statistical models have challenges in reproducing atomic-scale structures of nanoporous media.

Meanwhile, there has been a recent growth in advanced random noise models that open for new
ways to generate structured noise with statistical similarities to real samples, such as the use of Perlin noise. 
Such noise models have been used to model systems with complex statistical correlations such
as in GAMER\cite{groeneboom:2014}, and show great promise in being adaptable to many different physical systems. 
Also, such noise models are continuous and extendable, so that arbitrary large systems or
systems with arbitrary high precision can be modeled. 
While these models show great promise in several fields \cite{ebert2003texturing, cook2005wavelet, bridson2007curl, goldberg2008anisotropic},
they have not yet been systematically applied to characterize or generate nanoporous media with statistical properties corresponding to that of real systems.

@@

Another example, widely used in contemporary game content generation, is having prefabs, or
pre-fabricated 3D models such as doors, walls, tiles and roof segments automatically being
stitched together to create buildings.
When designing a city, instead of creating each building manually, you only need specify the
position, area and height of a structure.
Then, by using an algorithm that combines these prefabs to render the building, it is possible
to create cities of arbitrary size with a single command.
Naturally, it is also possible to create procedural methods that design the layout and placement of these buildings.

@@

An overview of the current state of procedural noise functions can be
found in \cite{lagae2010state}. Perlin noise is a band limited repeatable pseudo-random function
from $\mathbb R^n \to \mathbb R$ that is computationally fast, and has
properties which make it suitable for creating patterns that mimic
nature. Perlin noise is an approximation to Gaussian filtered noise
implemented as a pseudo-random spline. The algorithm uses a predefined
grid of gradients pointing in a random direction, where for any point
within a grid, we interpolate the gradients from the four corners. The
algorithm is as follows:
\begin{itemize}
  \item[1.] For a point, obtain the four closest gradients in the grid.
   \item [2.] For each of the gradients, calculate the dot product
     between the gradient and a vector defined by the  the distance
     between the the point and the current grid corner. 
   \item [3.] Instead of linear interpolation $F(t) = t$, use a fade function to produce
   smoother interpolated values. Perlin noise typically uses the following fade function: $F(t) = t ^3 (t  (6 - 15) + 10)$.
    \item [4.] Obtain the final value by linearly interpolating between the x-axis and
      use this result to linearly interpolate the y-axis. 
\end{itemize} 
Simplex noise works in a similar manner, but works on on simplices instead of hypergrids,
and so requires fewer computational steps, scales better in higher ($\ge 3$) dimensions
and has fewer anisotropic properties.

@@

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
$N$-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\Omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\Omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\Omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In one dimension, we show equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves
in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern is depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better
with higher dimensions and produces less asymmetric effects, we choose to use
simplex noise over regular Perlin noise as the main noise function throughout this paper.

@@

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce
nanoporous structures that are statistically similar to physical simulations of porous media.
For a given data set and a model, we wish to estimate the parameters that best represent the data. 
This is performed through a full likelihood analysis in order to obtain the posterior
probability distribution of the parameters. 
The model measure of the analysis is the radial distribution function $g(r)$, for
which the parameters will be adjusted to maximise.
In addition, we also investigate the porosity and surface area of the system. 
The simulated system is confined to a cube of size $\SI{159} {\angstrom}^3$, but since
the procedural noise function is continuous on all scales, the box could be of arbitrary size.

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image:
low persistence (large scale dominates), middle image: medium persistence (small
scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}
We construct a generic sum over procedural noise octaves with the expression: 
\begin{equation}
  M_1(\bar \theta) = M_1(\vec x, \psi, \Omega, \nu) = \sum_{\omega=0}^{\Omega} \psi^{\omega -1}   N\big((\vec x + \vec\sigma \cdot \omega)\nu \cdot 2^{\omega-1} \big),
\label{eq:noisemodel1}
\end{equation}
where $\Omega$ is the number of octaves, $\psi$ the persistence (scale falloff),
$N$ a generic $N$-dimensional noise function, $\vec \sigma$ a predefined random
vector that removes anisotropic artefacts and $\nu$ the frequency modifier representing
the initial largest scales in the model. For simplicity, $\bar \theta$ here denotes
the set of model parameters. For each mode, the frequency is thus multiplied, while
the overall amplitude is damped by a given persistence. In the end, we simulate a
large set of bulk particles without any walls, and use the noise function to remove
particles that are not within a procedural void given by a threshold $\tau$. Three
examples of this model are depicted in figure \ref{fig:model_example}, where we have
used $\Omega=6$ octaves and the scale is $\nu=0.01$ and cut-off $\tau=0$. The left
image depicts low persistence ($\psi = 0.0$), where only the first octave dominates
(corresponding to large scales). The middle picture shows the same parameters with
a higher persistence $(\psi = 0.5)$, where smaller scales become more dominating.
In the rightmost picture, the effect of the cut-off threshold $\tau = -0.3$ is shown to emphasize threshold cut-off.

@@

\begin{figure*}
\includegraphics[width=.99\textwidth]{mock_data_results_pts.png}
\caption{Results from the model validation simulation.The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Note that all input parameters were reproduced well within $1 \sigma$ (inner contour line).}
\label{fig:mockdataresults_pts}
\end{figure*}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{gofr_figure.pdf}
\caption{$g(r)$ for three data sets: the simulated SiO$_2$ input data set (red), the best-fit parameter model (blue) and a random low-fit parameter model (green).}
\label{fig:gofr1}
\end{figure}
Finally, we depict a comparison of the actual simulated input SiO$_2$ data set together with a
representation of the best-fit parameters in figure \ref{fig:porous_vs_model}   
\begin{figure}
\includegraphics[width=.45\textwidth]{comparison.png}
\caption{Right side: A realisation of the best-fit parameters $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$.Left side: Simulated SiO$_2$ input data set used for parameter estimation. Although the pore geometry seems to be similar, we see some artifacts, such as single loose atoms in the generated geometry.}
\label{fig:porous_vs_model}
\end{figure}
